Stack Safety as a Security Property

### 2020-07-10 main.tex

In this section we examine how an existing enforcement mechanism,
Depth Isolation from \citet{DBLP:conf/sp/RoesslerD18}, can implement
the formal stack safety property described in the previous section: it in
fact enforces a stronger, inductive version of that property which strengthens the
local state encapsulation component by checking for integrity and
confidentiality violations at every step of the callee's execution trace.

%\paragraph*{Micropolicies}
%
The enforcement policy of \citeauthor{DBLP:conf/sp/RoesslerD18} relies
on a programmable, tag-based reference monitor that runs alongside the
program.  To control this monitor, they use a programming model that
allows fine-grained manipulation of metadata tags to encode so-called
\emph{micropolicies}~\citep{pump_oakland2015}.
In such a system, all values in
memory addresses and registers (including the $\PCname$)
are enriched with an abstract metadata tag, which can represent 
arbitrary information about the value. A micropolicy is defined as a set of tags and a
collection of software-defined rules, indexed by machine opcode. 
At each step of the machine, the relevant rule is applied to the tags on 
the instructions inputs (PC, registers, memory) and on the instruction itself, and produces
one of two outcomes: either the instruction is allowed to execute (and generates
tags for the result of the operation and the new PC), or the machine
fail-stops with a policy violation. 
\citeauthor{pump_oakland2015} have shown that a wide range of micropolicies can
be defined using this scheme.

Efficient execution of these micropolicies relies on hardware implementations,
such as the PUMP architecture~\citep{pump:asplos2015}.  Tags are represented
as word-size bit vectors, stored separately from the regular memory and registers.
The hardware incorporates a rule cache
to allow quick retrieval of rule outputs for mapped inputs. If the cache misses,
the hardware traps to a software handler (running in a privileged context or
on a co-processor) to compute the rule result. To obtain adequate performance,
it is important to design micropolicies so that they hit in the cache as
much as possible. Thus, practical policies maintain a small working set of
distinct tags.
%% %
%% \rb{Maybe go into more detail, examples later\ldots or try to segue into stack
%% policies}

\subsection{A Conservative Policy}
\label{sec:conservative}
%
The first micropolicy we present to enforce the definition of stack
safety developed in \cref{sec:lse-and-wbcf} is a small variation
on the Depth Isolation policy presented by
\citet{DBLP:conf/sp/RoesslerD18}. Their policy tags the stack memory
with ownership information associated to each stack frame (and to each
separate object inside that frame, a more finely grained access
control that we do not need to consider here), and tags registers containing
stack pointers with access permission information. The compiler also
uses tags on instructions to identify the code 
sequences that have permission to manage the stack, say during calls and returns.
%% \rb{Conceivably on alloc
%%   operations, etc., which we have not yet mentioned.}

Our micropolicy works as follows. (We assume for ease of exposition that no
arguments are passed on the stack.)
%% \rb{Come back to this later?}.
The micropolicy maintains tags on values in memory of the form $\tagStackDepth{n}$,
indicating locations that belong to the stack frame at activation depth $n$,
or $\tagNoDepth$; it tags the $\PCname$ with tags of the form $\tagPCDepth{n}$.
During normal execution, the micropolicy rules only permit load and store operations
when the target memory is tagged with the same depth as the current {\PCname} tag.
Initially, the entire stack is tagged $\tagNoDepth$,
%\rb{actually, the ``initial frame'' should be tagged with its proper depth}
and the {\PCname} has tag $\tagPCDepth{0}$. These tags are altered
at exactly those points in
the program where the contour of the stack safety property changes:

@@

\item From caller to callee, when the machine executes an instruction marked as a call in
  the call map. 
  At this point, the current
  $\tagPCDepth{n}$ tag is incremented to $\tagPCDepth{(n + 1)}$, and the function
  entry sequence initializes  all locations of the new function frame with
  tag $\tagStackDepth{(n + 1)}$.

\item From callee back to caller, when execution is about to reach the return point
  $\ret{c}$
  corresponding to the call at $c$.  The exit sequence retags all locations in the function
  frame with $\tagNoDepth$. At the actual return instruction, the
  $\PCname$ tag is decremented.

@@

This discipline suffices to enforce local stack encapsulation.
To implement it,  we define ``blessed'' instruction sequences
intended to appear at the entry and exit of each function,
which manipulate tags as just described in addition to performing the
usual calling convention tasks of saving/restoring the return address to/from
the stack and adjusting the stack pointer. With the aid of an additional tag on
the $\rsp$ register, these sequences also serve to enforce
well-bracketed control flow.
The micropolicy guarantees atomic execution of these sequences
using a combination of tags on the instructions
and an additional tag on the $\PCname$; we omit the details here.

There remains the question of how to ensure that the sequences are
invoked at the right places. For a program to enjoy the stack safety property
(with respect to a particular call map), all we need require is that 
the entry sequence be initiated at any instruction marked as a call
in the call map. This is achieved by giving these instructions (another) special tag,
and it is easy to check statically that this has been done correctly.
If the code fails to initiate an exit
sequence at a point where the stack safety property expects a return,
the micropolicy will incorrectly behave as if execution
is continuing in the callee, but since the callee never has stronger access
rights than the caller, this is harmless.

%% APT: removed because the tag part is too mysterious and the instruction part
%% is standard. 
%% This is the entry
%% sequence:
%% %
%% \setcounter{pcctr}{1}
%% \[
%%   \begin{array}{l|l|l|l}
%%      & \mathit{Instruction} & \mathit{Tags} & \\
%%     \hline
%%     \row{\sw ~ \rsp ~ \tta ~ 1}{[\rsp+1] \leftarrow \rra}{\tagHa, \tagInstr}
%%         {Store return address in stack}
%%     \row{\addi ~ \rsp ~ \rsp ~ 2}{\rsp \leftarrow \rsp + 2}{\tagHb, \tagInstr}
%%         {Increment stack pointer by frame size}
%%   \end{array}
%% \]
%% %
%% And the exit sequence:
%% %
%% \setcounter{pcctr}{1}
%% \[
%%   \begin{array}{l|l|l|l}
%%      & \mathit{Instruction} & \mathit{Tags} & \\
%%     \hline
%%     \row{\lw ~ \rra ~ \rsp ~ \negate 1}{\rra \leftarrow [\rsp - 1]}{\tagRa, \tagInstr}
%%         {Load return address}
%%     \row{\addi ~ \rsp ~ \rsp ~ \negate 2}{\rsp \leftarrow \rsp - 2}{\tagRb, \tagInstr}
%%         {Decrement stack pointer by frame size}
%%     \row{\jalr ~ \rra ~ \rra ~ 0\apt{wrong}}{\PCname \leftarrow \rra}{\tagRc, \tagInstr}
%%         {Return to caller ($\jalr$)}
%%   \end{array}
%% \]

@@

%% These sequences also enforce the $\SP$ discipline. When stack frames
%% are of fixed size, it suffices to increment $\SP$ and tag it with a
%% dedicated tag in the header sequence, which is later checked after
%% decrementing $\SP$ during the exit sequence to authorize the
%% return. The well-formedness of the blessed sequences can be checked
%% statically.

%% Only programs whose sequences are well-formed are
%% protected by the micropolicy, although programs are still protected
%% (by fail-stopping) even if the intended sequences are missing. By
%% tracking the execution of the blessed sequences in the tag state of
%% the {\PCname} register, the micropolicy ensures that the entry
%% sequences of calls declared in the call map are executed correctly.

Note that in this policy both the entry and exit sequences
must write to each element of
the frame, which can be quite expensive, especially for programs that
allocate large but sparsely populated frames; we return to this point
in \cref{sec:lazy}.
%
%\rb{Note that we have also considered and implemented more flexible schemes
%  based on allowing operations above the current activation depth. How do we put
%  these together?}
%\rb{Do we want to discuss variations involving frame pointers or more implicit
%  return addresses, argument passing on the stack, alloc/dealloc, etc.?}
%\rb{Where to talk about the role of the compiler?}

>>>>

In this section we are going to see how an existing enforcement mechanism,
Depth Isolation from \citet{DBLP:conf/sp/RoesslerD18}, can implement
the formal stack safety property described in the previous section: it in
fact enforces a stronger, inductive version of it that strengthens the
local state encapsulation component by checking for integrity and
confidentiality violations at every step of the callee's execution trace.

\paragraph*{Micropolicies}
%
The enforcement policy of \citeauthor{DBLP:conf/sp/RoesslerD18} relies
on a programmable, tag-based reference monitor that runs alongside the
program.  To control this monitor, they use a programming model that
allows fine-grained manipulation of metadata tags to encode so-called
\emph{micropolicies}~\citep{pump_oakland2015}.

In such a system, all memory addresses and registers (including the program
counter) are enriched with an abstract metadata tag, which can represent an
arbitrary data structure. A micropolicy is defined as a set of tags, and a
collection of software-defined rules that represent the transfer function of the
policy. Each rule inspects the tags associated to the inputs of an instruction
(PC, registers, memory) as well as the instruction opcode itself, and produces
one of two outcomes: either the instruction is allowed to execute (and generates
and assigns tags for the new PC and the result of the operation), or the machine
fail-stops with a policy violation. During program execution, micropolicy rules
are checked before each machine step.

Efficient execution of these micropolicies relies on hardware implementations,
such as the PUMP architecture. This hardware incorporates a monitor rule cache
to allow quick retrieval of transfer function outputs for mapped inputs. Cache
misses are detected and trapped to software handlers that use the software rules
of the policy to compute the relevant parts of the transfer function.
\citeauthor{pump_oakland2015} have shown that a wide range of micropolicies can
be defined and enforced efficiently in these machines.
%% %
%% \rb{Maybe go into more detail, examples later\ldots or try to segue into stack
%% policies}

\subsection{A Conservative Policy}
\label{sec:conservative}
%
The first micropolicy we present to enforce the definition of stack
safety developed in \cref{sec:lse-and-wbcf} is a small variation
on the Depth Isolation policy presented by
\citet{DBLP:conf/sp/RoesslerD18}. Their policy tags the stack memory
with ownership information associated to each stack frame (and to each
separate object inside that frame, a more finely grained access
control that we do not need to consider here), and tags registers containing
stack pointers with access permission information. Beyond
those capabilities, the compiler generates designated sequences of
instructions with special tags that grant them permission to manage
the stack, say during calls and returns.
%% \rb{Conceivably on alloc
%%   operations, etc., which we have not yet mentioned.}

Our micropolicy works as follows. We assume for ease of exposition that no
arguments are passed on the stack.
%% \rb{Come back to this later?}.
Initially, the entire stack is tagged as unused with an $\tagNoDepth$ tag,
%\rb{actually, the ``initial frame'' should be tagged with its proper depth}
and the {\PCname} carries in its tag the depth of the current
activation at all times, starting with a $\tagPCDepth{0}$ tag. The
current activation depth needs to change exactly at those points in
the program where the contour of the stack safety property changes:

@@

\item From caller to callee, when the machine executes an instruction tagged
  with $\tagCall$ as belonging to the call map. At this point, the current
  {\PCname} depth $n$ is incremented to $\tagPCDepth{n + 1}$; on function entry,
  the locations of the newly allocated frame are marked $\tagStackDepth{n + 1}$.

\item From callee back to caller, when execution reaches the corresponding
  return instruction for the call instruction, as defined by
  $\ret{\mach_c}\ \mach_r$. At this point, the {\PCname} tag is decremented.
  Prior to this point, the locations of the old frame have been restored to
  $\tagNoDepth$.

@@

To enforce this discipline, we define blessed header and return instruction
sequences. These are specific sequences of tagged instructions that are intended
to appear at the beginning and end of each function. This is the header
sequence:
%
\setcounter{pcctr}{1}
\[
  \begin{array}{l|l|l|l}
     & \mathit{Instruction} & \mathit{Tags} & \\
    \hline
    \row{\sw ~ \rsp ~ \tta ~ 1}{[\rsp+1] \leftarrow \rra}{\tagHa, \tagInstr}
        {Store return address in stack}
    \row{\addi ~ \rsp ~ \rsp ~ 2}{\rsp \leftarrow \rsp + 2}{\tagHb, \tagInstr}
        {Increment stack pointer by frame size}
  \end{array}
\]
%
And the return sequence:
%
\setcounter{pcctr}{1}
\[
  \begin{array}{l|l|l|l}
     & \mathit{Instruction} & \mathit{Tags} & \\
    \hline
    \row{\lw ~ \rra ~ \rsp ~ \negate 1}{\rra \leftarrow [\rsp - 1]}{\tagRa, \tagInstr}
        {Load return address}
    \row{\addi ~ \rsp ~ \rsp ~ \negate 2}{\rsp \leftarrow \rsp - 2}{\tagRb, \tagInstr}
        {Decrement stack pointer by frame size}
    \row{\jalr ~ \rra ~ \rra ~ 0\apt{wrong}}{\PCname \leftarrow \rra}{\tagRc, \tagInstr}
        {Return to caller ($\jalr$)}
  \end{array}
\]

@@

These sequences also enforce the $\SP$ discipline. When stack frames
are of fixed size, it suffices to increment $\SP$ and tag it with a
dedicated tag in the header sequence, which is later checked after
decrementing $\SP$ during the exit sequence to authorize the
return. The well-formedness of the blessed sequences can be checked
statically. Only programs whose sequences are well-formed are
protected by the micropolicy, although programs are still protected
(by fail-stopping) even if the intended sequences are missing. By
tracking the execution of the blessed sequences in the tag state of
the {\PCname} register, the micropolicy ensures that the entry
sequences of calls declared in the call map are executed correctly.
%
During normal execution, the {\PCname} uses the current activation
depth in its tag $\tagPCDepth{n}$ to gate load and store operations,
which are allowed to operate only on the corresponding stack frame,
tagged $\tagStackDepth{n}$.
%\rb{Note that we have also considered and implemented more flexible schemes
%  based on allowing operations above the current activation depth. How do we put
%  these together?}
%\rb{Do we want to discuss variations involving frame pointers or more implicit
%  return addresses, argument passing on the stack, alloc/dealloc, etc.?}
%\rb{Where to talk about the role of the compiler?}
